---
title: "GRAT-Index"
author: "Florida Community Innovation GRAT Team"
date: "2025-1-20"
output: html_document
---
```{r setup, include = F, message = F}
library(tidycensus)
library(tidyverse)
library(dplyr)
library(cluster)
library(ggplot2)
census_api_key("b04a3c521c0b4e5c1a74610ea5c18a5d31ed9aac")
```

## Census Data
```{r, message = F, warning = F, output = F}
# Store variables here
variables <- c("B25001_001", "B25004_001", "B25058_001", "B25077_001", "B25003_003","B25003_002", "B03002_003", "B03002_004", "B03002_001", "B19013_001","B17001_002", "B17001_001", "B25070_007", "B25070_008", "B25070_009","B25070_010", "B25070_011", "B25091_009", "B25091_010", "B25091_011","B25091_012", "B25091_013", "B25095_005", "B25095_006", "B25095_007","B25095_008"
)

# Get census data for our three time periods (will adjust after MVP)
census_data_2020 <- get_acs(
  geography = "tract",
  state = "FL",
  county = "Alachua",
  variables = variables,
  geometry = FALSE,
  year = 2020
)

census_data_2022 <- get_acs(
  geography = "tract",
  state = "FL",
  county = "Alachua",
  variables = variables,
  geometry = FALSE,
  year = 2022
)

census_data_2023 <- get_acs(
  geography = "tract",
  state = "FL",
  county = "Alachua",
  variables = variables,
  geometry = FALSE,
  year = 2023
)
```

```{r}
census_data_2020 <- census_data_2020|>
  pivot_wider(
    id_cols = c(GEOID, NAME),
    names_from = variable,
    values_from = estimate
  ) |> group_by(GEOID)

census_data_2022 <- census_data_2022|>
  pivot_wider(
    id_cols = c(GEOID, NAME),
    names_from = variable,
    values_from = estimate
  ) |> group_by(GEOID)

census_data_2023 <- census_data_2023|>
  pivot_wider(
    id_cols = c(GEOID, NAME),
    names_from = variable,
    values_from = estimate
    ) |> group_by(GEOID)
```

## Data processing

``` {r}
# Need to fix census_data_2020 NAME column delimiter
library(stringr)

census_data_2020 <- census_data_2020 %>%
  mutate(NAME = str_replace_all(NAME, ",", ";"))
```

```{r}
census_data_2020 <- census_data_2020 %>%
  rename(
    Median_Contract_Rent = B25058_001,
    Median_Home_Value = B25077_001,
    Total_Renter_Occupied_Units = B25003_003,
    Vacancy_Rate = B25004_001,
    Total_Owner_Occupied_Units = B25003_002,
    White_Alone = B03002_003,
    Black_Alone = B03002_004,
    Total_Population = B03002_001,
    Median_Household_Income = B19013_001,
    Total_Housing_Units = B25001_001,
    Below_Poverty_Level = B17001_002,
    Total_Poverty_Population = B17001_001,
    Renter_Paying_30_34 = B25070_007,
    Renter_Paying_35_39 = B25070_008,
    Renter_Paying_40_49 = B25070_009,
    Renter_Paying_50_59 = B25070_010,
    Renter_Paying_60_or_More = B25070_011,
    Owner_With_Mortgage_Paying_30_34 = B25091_009,
    Owner_With_Mortgage_Paying_35_39 = B25091_010,
    Owner_With_Mortgage_Paying_40_49 = B25091_011,
    Owner_With_Mortgage_Paying_50_59 = B25091_012,
    Owner_With_Mortgage_Paying_60_or_More = B25091_013,
    Owner_Without_Mortgage_Paying_30_34 = B25095_005,
    Owner_Without_Mortgage_Paying_35_39 = B25095_006,
    Owner_Without_Mortgage_Paying_40_49 = B25095_007,
    Owner_Without_Mortgage_Paying_50_or_More = B25095_008
  )

census_data_2022 <- census_data_2022 %>%
  rename(
    Total_Housing_Units = B25001_001,
    Median_Contract_Rent = B25058_001,
    Median_Home_Value = B25077_001,
    Total_Renter_Occupied_Units = B25003_003,
    Vacancy_Rate = B25004_001,
    Total_Owner_Occupied_Units = B25003_002,
    White_Alone = B03002_003,
    Black_Alone = B03002_004,
    Total_Population = B03002_001,
    Median_Household_Income = B19013_001,
    Total_Housing_Units = B25001_001,
    Below_Poverty_Level = B17001_002,
    Total_Poverty_Population = B17001_001,
    Renter_Paying_30_34 = B25070_007,
    Renter_Paying_35_39 = B25070_008,
    Renter_Paying_40_49 = B25070_009,
    Renter_Paying_50_59 = B25070_010,
    Renter_Paying_60_or_More = B25070_011,
    Owner_With_Mortgage_Paying_30_34 = B25091_009,
    Owner_With_Mortgage_Paying_35_39 = B25091_010,
    Owner_With_Mortgage_Paying_40_49 = B25091_011,
    Owner_With_Mortgage_Paying_50_59 = B25091_012,
    Owner_With_Mortgage_Paying_60_or_More = B25091_013,
    Owner_Without_Mortgage_Paying_30_34 = B25095_005,
    Owner_Without_Mortgage_Paying_35_39 = B25095_006,
    Owner_Without_Mortgage_Paying_40_49 = B25095_007,
    Owner_Without_Mortgage_Paying_50_or_More = B25095_008
  )

census_data_2023 <- census_data_2023 %>%
  rename(
    Median_Contract_Rent = B25058_001,
    Median_Home_Value = B25077_001,
    Total_Renter_Occupied_Units = B25003_003,
    Vacancy_Rate = B25004_001,
    Total_Owner_Occupied_Units = B25003_002,
    White_Alone = B03002_003,
    Black_Alone = B03002_004,
    Total_Population = B03002_001,
    Median_Household_Income = B19013_001,
    Total_Housing_Units = B25001_001,
    Below_Poverty_Level = B17001_002,
    Total_Poverty_Population = B17001_001,
    Renter_Paying_30_34 = B25070_007,
    Renter_Paying_35_39 = B25070_008,
    Renter_Paying_40_49 = B25070_009,
    Renter_Paying_50_59 = B25070_010,
    Renter_Paying_60_or_More = B25070_011,
    Owner_With_Mortgage_Paying_30_34 = B25091_009,
    Owner_With_Mortgage_Paying_35_39 = B25091_010,
    Owner_With_Mortgage_Paying_40_49 = B25091_011,
    Owner_With_Mortgage_Paying_50_59 = B25091_012,
    Owner_With_Mortgage_Paying_60_or_More = B25091_013,
    Owner_Without_Mortgage_Paying_30_34 = B25095_005,
    Owner_Without_Mortgage_Paying_35_39 = B25095_006,
    Owner_Without_Mortgage_Paying_40_49 = B25095_007,
    Owner_Without_Mortgage_Paying_50_or_More = B25095_008
  )
```


```{r}
census_data_2020 <- census_data_2020 %>%
  mutate(
    Vacancy_Rate = (Total_Housing_Units - Total_Renter_Occupied_Units - Total_Owner_Occupied_Units) / Total_Housing_Units * 100,
    Percent_White_Alone = White_Alone / Total_Population * 100,
    Percent_Black_Alone = Black_Alone / Total_Population * 100,
    Poverty_Rate = Below_Poverty_Level / Total_Poverty_Population * 100,
    Total_Renter_Paying_30 = Renter_Paying_30_34 + Renter_Paying_35_39 + 
                             Renter_Paying_40_49 + Renter_Paying_50_59 + 
                             Renter_Paying_60_or_More,
    Percent_Renter_Paying_30 = (Total_Renter_Paying_30 / Total_Renter_Occupied_Units) * 100,
    
    Total_Owner_With_Mortgage_Paying_30 = Owner_With_Mortgage_Paying_30_34 + 
                                          Owner_With_Mortgage_Paying_35_39 + 
                                          Owner_With_Mortgage_Paying_40_49 + 
                                          Owner_With_Mortgage_Paying_50_59 + 
                                          Owner_With_Mortgage_Paying_60_or_More,
    Percent_Owner_With_Mortgage_Paying_30 = (Total_Owner_With_Mortgage_Paying_30 / Total_Owner_Occupied_Units) * 100,
        Total_Owner_Without_Mortgage_Paying_30 = Owner_Without_Mortgage_Paying_30_34 + 
                                             Owner_Without_Mortgage_Paying_35_39 + 
                                             Owner_Without_Mortgage_Paying_40_49 + 
                                             Owner_Without_Mortgage_Paying_50_or_More,
    Percent_Owner_Without_Mortgage_Paying_30 = (Total_Owner_Without_Mortgage_Paying_30 / Total_Owner_Occupied_Units) * 100,
    
    Vacancy_Rate_Percent = (Vacancy_Rate / Total_Housing_Units) * 100,
    
    Percent_Below_Poverty = (Below_Poverty_Level / Total_Poverty_Population) * 100 
  )
census_data_2022 <- census_data_2022 %>%
  mutate(
     Vacancy_Rate = (Total_Housing_Units - Total_Renter_Occupied_Units - Total_Owner_Occupied_Units) / Total_Housing_Units * 100,
    Percent_White_Alone = White_Alone / Total_Population * 100,
    Percent_Black_Alone = Black_Alone / Total_Population * 100,
    Poverty_Rate = Below_Poverty_Level / Total_Poverty_Population * 100,
    Total_Renter_Paying_30 = Renter_Paying_30_34 + Renter_Paying_35_39 + 
                             Renter_Paying_40_49 + Renter_Paying_50_59 + 
                             Renter_Paying_60_or_More,
    Percent_Renter_Paying_30 = (Total_Renter_Paying_30 / Total_Renter_Occupied_Units) * 100,
    
    Total_Owner_With_Mortgage_Paying_30 = Owner_With_Mortgage_Paying_30_34 + 
                                          Owner_With_Mortgage_Paying_35_39 + 
                                          Owner_With_Mortgage_Paying_40_49 + 
                                          Owner_With_Mortgage_Paying_50_59 + 
                                          Owner_With_Mortgage_Paying_60_or_More,
    Percent_Owner_With_Mortgage_Paying_30 = (Total_Owner_With_Mortgage_Paying_30 / Total_Owner_Occupied_Units) * 100,
        Total_Owner_Without_Mortgage_Paying_30 = Owner_Without_Mortgage_Paying_30_34 + 
                                             Owner_Without_Mortgage_Paying_35_39 + 
                                             Owner_Without_Mortgage_Paying_40_49 + 
                                             Owner_Without_Mortgage_Paying_50_or_More,
    Percent_Owner_Without_Mortgage_Paying_30 = (Total_Owner_Without_Mortgage_Paying_30 / Total_Owner_Occupied_Units) * 100,
    
    Vacancy_Rate_Percent = (Vacancy_Rate / Total_Housing_Units) * 100,
    
    Percent_Below_Poverty = (Below_Poverty_Level / Total_Poverty_Population) * 100 
  )

census_data_2023 <- census_data_2023 %>%
  mutate(
     Vacancy_Rate = (Total_Housing_Units - Total_Renter_Occupied_Units - Total_Owner_Occupied_Units) / Total_Housing_Units * 100,
    Percent_White_Alone = White_Alone / Total_Population * 100,
    Percent_Black_Alone = Black_Alone / Total_Population * 100,
    Poverty_Rate = Below_Poverty_Level / Total_Poverty_Population * 100,
    Total_Renter_Paying_30 = Renter_Paying_30_34 + Renter_Paying_35_39 + 
                             Renter_Paying_40_49 + Renter_Paying_50_59 + 
                             Renter_Paying_60_or_More,
    Percent_Renter_Paying_30 = (Total_Renter_Paying_30 / Total_Renter_Occupied_Units) * 100,
    
    Total_Owner_With_Mortgage_Paying_30 = Owner_With_Mortgage_Paying_30_34 + 
                                          Owner_With_Mortgage_Paying_35_39 + 
                                          Owner_With_Mortgage_Paying_40_49 + 
                                          Owner_With_Mortgage_Paying_50_59 + 
                                          Owner_With_Mortgage_Paying_60_or_More,
    Percent_Owner_With_Mortgage_Paying_30 = (Total_Owner_With_Mortgage_Paying_30 / Total_Owner_Occupied_Units) * 100,
        Total_Owner_Without_Mortgage_Paying_30 = Owner_Without_Mortgage_Paying_30_34 + 
                                             Owner_Without_Mortgage_Paying_35_39 + 
                                             Owner_Without_Mortgage_Paying_40_49 + 
                                             Owner_Without_Mortgage_Paying_50_or_More,
    Percent_Owner_Without_Mortgage_Paying_30 = (Total_Owner_Without_Mortgage_Paying_30 / Total_Owner_Occupied_Units) * 100,
    
    Vacancy_Rate_Percent = (Vacancy_Rate / Total_Housing_Units) * 100,
    
    Percent_Below_Poverty = (Below_Poverty_Level / Total_Poverty_Population) * 100 
  )
```

## Replace Median Home Price Data with census api data?
```{r}
vars <- c(median_home_value = "B25077_001E")

home_prices_2020 <- get_acs(geography = "tract", variables = vars, 
                            state = "FL", year = 2020, survey = "acs5")

home_prices_2022 <- get_acs(geography = "tract", variables = vars, 
                            state = "FL", year = 2022, survey = "acs5")

home_prices_2023 <- get_acs(geography = "tract", variables = vars, 
                            state = "FL", year = 2023, survey = "acs5")

home_prices <- home_prices_2020 %>%
  rename(median_home_value_2020 = estimate) %>%
  dplyr::select(GEOID, NAME, median_home_value_2020) %>%
  left_join(home_prices_2022 %>% rename(median_home_value_2022 = estimate) %>% dplyr::select(GEOID, median_home_value_2022), by = "GEOID") %>%
  left_join(home_prices_2023 %>% rename(median_home_value_2023 = estimate) %>% dplyr::select(GEOID, median_home_value_2023), by = "GEOID")

# Compute percentage changes
home_prices <- home_prices %>%
  mutate(
    change_2020_2022 = (median_home_value_2022 - median_home_value_2020) / median_home_value_2020 * 100,
    change_2022_2023 = (median_home_value_2023 - median_home_value_2022) / median_home_value_2022 * 100,
    change_2020_2023 = (median_home_value_2023 - median_home_value_2020) / median_home_value_2020 * 100
  ) %>%
  drop_na()
```

```{r}
state_avg <- home_prices %>%
  pivot_longer(cols = starts_with("median_home_value_"), 
               names_to = "year", 
               values_to = "median_home_price") %>%
  mutate(year = gsub("median_home_value_", "", year)) %>%  
  group_by(year) %>%
  summarize(state_median_price = mean(median_home_price, na.rm = TRUE)) %>%
  pivot_wider(names_from = year, values_from = state_median_price) %>%
  mutate(
    change_2020_2022 = (`2022` - `2020`) / `2020` * 100,
    change_2022_2023 = (`2023` - `2022`) / `2022` * 100,
    change_2020_2023 = (`2023` - `2020`) / `2020` * 100
  )

tract_changes <- home_prices %>%
  mutate(
    tract_change_2020_2022 = (median_home_value_2022 - median_home_value_2020) / median_home_value_2020 * 100,
    tract_change_2022_2023 = (median_home_value_2023 - median_home_value_2022) / median_home_value_2022 * 100,
    tract_change_2020_2023 = (median_home_value_2023 - median_home_value_2020) / median_home_value_2020 * 100
  ) %>%
  mutate(
    over_under_2020_2022 = tract_change_2020_2022 - state_avg$change_2020_2022,
    over_under_2022_2023 = tract_change_2022_2023 - state_avg$change_2022_2023,
    over_under_2020_2023 = tract_change_2020_2023 - state_avg$change_2020_2023
  ) |>
  drop_na()

tract_changes <- tract_changes %>%
  rename(
    median_home_value_2020_tract = median_home_value_2020,
    median_home_value_2022_tract = median_home_value_2022,
    median_home_value_2023_tract = median_home_value_2023,
    change_2020_2022_tract = change_2020_2022,
    change_2022_2023_tract = change_2022_2023,
    change_2020_2023_tract = change_2020_2023
  )

tract_changes <- tract_changes %>%
  mutate(NAME = str_replace_all(NAME, ",", ";"))
```

```{r}
append_year_suffix <- function(data, year) {
  suffix <- paste0("_", year)
  data %>%
    rename_with(~ paste0(.x, suffix), -c(GEOID, NAME))
}

census_data_2023 <- append_year_suffix(census_data_2023, "2023")
census_data_2022 <- append_year_suffix(census_data_2022, "2022")
census_data_2020 <- append_year_suffix(census_data_2020, "2020")

combined_census_data <- census_data_2023 %>%
  full_join(census_data_2022, by = c("GEOID", "NAME")) %>%
  full_join(census_data_2020, by = c("GEOID", "NAME"))

combined_census_data <- combined_census_data %>% distinct(GEOID, .keep_all = TRUE)
```


```{r}
# calculate change (2023 - 2020) for each variable
combined_data_test <- combined_census_data %>%
  mutate(across(ends_with("2020"), 
                ~ as.numeric(pick(str_replace(cur_column(), "2020$", "2023"))) - ., 
                .names = "change_{.col}"), 
         .keep = "all")
```

``` {r}
change_data <- combined_data_test %>%
  mutate(RegionName = str_extract(NAME, "^[^,]+")) %>%  # tract-level data
  distinct(GEOID, .keep_all = TRUE)

colnames(change_data)

# replace null values with mean of column
change_data[] <- lapply(change_data, function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x))
```

## Kmeans Clustering to Classify Gentrification
```{r}
change_data_kmeans <- data.frame(change_data)
kmeans_data <- change_data_kmeans %>%
  dplyr::select(Total_Poverty_Population_2023,
                Below_Poverty_Level_2023,
                Median_Household_Income_2020,
                Median_Household_Income_2023,
                Median_Home_Value_2023,
                Percent_White_Alone_2023,
                Percent_Black_Alone_2023,
                change_Total_Population_2020,
                change_Percent_White_Alone_2020,
                change_Percent_Black_Alone_2020,
                change_Poverty_Rate_2020,
                change_Total_Housing_Units_2020,
                change_Median_Household_Income_2020,
                change_Median_Home_Value_2020,
                change_Vacancy_Rate_Percent_2020
                )

kmeans_model <- kmeans(scale(kmeans_data), centers = 3)
change_data_kmeans$Cluster <- kmeans_model$cluster
```

```{r}
write.csv(change_data_kmeans, "C:\\Users\\szhan\\Documents\\FCI\\GRAT\\census_data_kmeans.csv", row.names=FALSE)
```


``` {r}
kmeans_centers <- kmeans_model$center
write.csv(kmeans_centers, "C:\\Users\\szhan\\Documents\\FCI\\GRAT\\census_data_kmeans_center.csv", row.names=FALSE)
as.data.frame(kmeans_model$center)
```

``` {r}
# pca_result <- prcomp(scale(kmeans_data))
# 
# plot_data <- data.frame(
#   PC1 = pca_result$x[, 1],
#   PC2 = pca_result$x[, 2],
#   cluster = as.factor(kmeans_result$cluster)  
# )
# 
# ggplot(plot_data, aes(x = PC1, y = PC2, color = cluster)) +
#   geom_point(size = 2) +
#   labs(title = "PCA Performed on K-Means Clusters",
#        x = "Principal Component 1", y = "Principal Component 2") +
#   theme_minimal()
```

## PCA

```{r} 
combined_data_for_pca <- left_join(tract_changes, change_data, by = "GEOID")
combined_data_for_pca <- combined_data_for_pca %>%
  filter(str_detect(NAME.x, "Alachua"))  # Assuming NAME contains county info
```

```{r}
pca_explanatory_vars <- combined_data_for_pca %>%
  dplyr::select(starts_with("change_"))

pca_explanatory_vars <- pca_explanatory_vars %>%  
  mutate(across(everything(), ~ as.numeric(unlist(.))))

# Replace 4 NA values for change_median_contract_rent_2020 with median of the column

pca_explanatory_vars <- pca_explanatory_vars %>% 
  mutate(change_Median_Contract_Rent_2020 = ifelse(
    is.na(change_Median_Contract_Rent_2020), 
    median(change_Median_Contract_Rent_2020, na.rm = TRUE), 
    change_Median_Contract_Rent_2020
  ))

dependent_var <- combined_data_for_pca$over_under_2020_2023  

pca_result <- prcomp(pca_explanatory_vars, scale. = TRUE, rank = 15)

summary(pca_result)

library(ggplot2)

pca_scores <- as.data.frame(pca_result$x)
pca_scores$over_under <- dependent_var  

ggplot(pca_scores, aes(x = PC1, y = PC2, color = over_under)) +
  geom_point() +
  labs(title = "PCA of Housing Changes (2020-2023)",
       x = "Principal Component 1",
       y = "Principal Component 2",
       color = "Over/Under (%)") +
  theme_minimal()
```

```{r}
loadings <- pca_result$rotation
print(loadings)

pc1_contributions <- loadings[order(abs(loadings[, "PC1"]), decreasing = TRUE), "PC1"]
print(pc1_contributions)

pc2_contributions <- loadings[order(abs(loadings[, "PC2"]), decreasing = TRUE), "PC2"]
print(pc2_contributions)
```

## Neural Network 

```{r}
library(keras)

pca_explanatory_vars_scaled <- scale(pca_explanatory_vars[, -1]) 

pca_data <- data.frame(pca_explanatory_vars_scaled, 
                       over_under_2020_2023 = combined_data_for_pca$over_under_2020_2023)
```

```{r, message=F}

split_index <- sample(1:nrow(pca_data), 0.8 * nrow(pca_data))  
train_data <- pca_data[split_index, ]
test_data <- pca_data[-split_index, ]

train_x <- as.matrix(train_data[, -which(names(train_data) == "over_under_2020_2023")])  
train_y <- train_data$over_under_2020_2023

test_x <- as.matrix(test_data[, -which(names(test_data) == "over_under_2020_2023")])
test_y <- test_data$over_under_2020_2023

train_x <- scale(train_x)
test_x <- scale(test_x, center = attr(train_x, "scaled:center"), scale = attr(train_x, "scaled:scale"))

train_y <- scale(train_y)
test_y <- scale(test_y, center = attr(train_y, "scaled:center"), scale = attr(train_y, "scaled:scale"))

model <- keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = ncol(train_x)) %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dense(units = 1, activation = "linear")  

model %>% compile(
  optimizer = optimizer_adam(learning_rate = 0.005),
  loss = "mean_squared_error",
  metrics = c("mean_absolute_error")
)

history <- model %>% fit(
  x = train_x,
  y = train_y,
  epochs = 50,
  batch_size = 32,
  validation_split = 0.2,
  verbose = 1
)

evaluation <- model %>% evaluate(test_x, test_y)
cat("Test Loss:", evaluation[["loss"]], "\n")
cat("Test MAE:", evaluation[["mean_absolute_error"]], "\n")

predictions <- model %>% predict(test_x)
comparison <- data.frame(Actual = test_y, Predicted = predictions)
```

```{r}
head(comparison)
```

## Comparison of Outputs

```{r}
create_folds <- function(data, k = 5, seed = 42) {
  set.seed(seed)
  data <- data[sample(nrow(data)), ]
  fold_indices <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
  
  folds <- lapply(1:k, function(i) {
    which(fold_indices == i)
  })
  
  return(folds)
}

folds <- create_folds(pca_data, k = 5)
```

# K-fold cross validation PCA

```{r}
# Getting rid of error " cannot rescale a constant/zero column to unit variance"
pca_predictions <- numeric(nrow(pca_data))

for (i in seq_along(folds)) {
  test_indices <- folds[[i]]
  train_data <- pca_data[-test_indices, ]
  test_data  <- pca_data[test_indices, ]
  
  # Identify non-constant columns (excluding the target)
  feature_names <- names(train_data)[-ncol(train_data)]
  non_const <- sapply(train_data[, feature_names], function(x) var(x) > 0)
  valid_features <- feature_names[non_const]
  
  # Subset both training and test data to only those features
  train_features <- train_data[, valid_features, drop = FALSE]
  test_features  <- test_data[, valid_features, drop = FALSE]
  
  # Combine with the target variable
  train_data_mod <- cbind(train_features, over_under = train_data$over_under_2020_2023)
  test_data_mod  <- cbind(test_features, over_under = test_data$over_under_2020_2023)
  
  # Run PCA on training features (without target)
  pca_train <- prcomp(train_data_mod[, -ncol(train_data_mod)], scale. = TRUE)
  
  # Extract first two PCs and attach the target
  train_scores <- as.data.frame(pca_train$x[, 1:2])
  train_scores$over_under <- train_data_mod$over_under
  
  # Predict PC scores for test set
  test_scores <- predict(pca_train, newdata = test_data_mod[, -ncol(test_data_mod)])[, 1:2]
  test_scores <- as.data.frame(test_scores)
  test_scores$over_under <- test_data_mod$over_under
  
  # Fit linear model on training PC scores
  model <- lm(over_under ~ ., data = train_scores)
  pca_predictions[test_indices] <- predict(model, newdata = test_scores)
}

pca_predictions <- 100 * (pca_predictions - min(pca_predictions)) / 
                   (max(pca_predictions) - min(pca_predictions))

```

## PCA variables plot

```{r}
pca_scores_normalized <- as.data.frame(pca_scores)
pca_scores_normalized$PC1 <- 100 * (pca_scores_normalized$PC1 - min(pca_scores_normalized$PC1)) / 
  (max(pca_scores_normalized$PC1) - min(pca_scores_normalized$PC1))
pca_scores_normalized$PC2 <- 100 * (pca_scores_normalized$PC2 - min(pca_scores_normalized$PC2)) / 
  (max(pca_scores_normalized$PC2) - min(pca_scores_normalized$PC2))

library(ggplot2)
pca_scores_normalized$GentrificationStatus <- ifelse(pca_data$over_under_2020_2023 > 10, "Gentrified", "Non-Gentrified")

ggplot(pca_scores_normalized, aes(x = PC1, y = PC2, color = GentrificationStatus)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Gentrified" = "red", "Non-Gentrified" = "green")) +
  labs(
    title = "Scatter Plot of Normalized PCA Variables",
    x = "Principal Component 1 (Normalized)",
    y = "Principal Component 2 (Normalized)",
    color = "Gentrification Status"
  ) +
  theme_minimal()
```

## K-fold cross-valid for neural network

```{r}
library(keras)
library(tensorflow)

create_model <- function(input_shape) {
  k_clear_session()  
  model <- keras_model_sequential() %>%
    layer_dense(units = 64, activation = "relu", input_shape = input_shape) %>%
    layer_dropout(rate = 0.3) %>%
    layer_dense(units = 32, activation = "relu") %>%
    layer_dropout(rate = 0.3) %>%
    layer_dense(units = 1, activation = "linear")
  
  model %>% compile(
    optimizer = optimizer_adam(learning_rate = 0.001),
    loss = "mean_squared_error",
    metrics = c("mean_absolute_error")
  )
  return(model)
}

neural_net_predictions <- numeric(nrow(pca_data)) 

for (i in seq_along(folds)) {
  test_indices <- folds[[i]]
  train_data <- pca_data[-test_indices, ]
  test_data <- pca_data[test_indices, ]
  
  train_x <- scale(as.matrix(train_data[, -ncol(train_data)]))
  test_x <- scale(as.matrix(test_data[, -ncol(test_data)]), 
                  center = attr(train_x, "scaled:center"), 
                  scale = attr(train_x, "scaled:scale"))
  
  train_y <- matrix(train_data$over_under_2020_2023, ncol = 1)
  test_y <- matrix(test_data$over_under_2020_2023, ncol = 1)
  train_x <- array_reshape(train_x, dim = c(nrow(train_x), ncol(train_x)))
  test_x <- array_reshape(test_x, dim = c(nrow(test_x), ncol(test_x)))

  train_x <- tf$convert_to_tensor(train_x, dtype = tf$float32)
  test_x <- tf$convert_to_tensor(test_x, dtype = tf$float32)
  train_y <- tf$convert_to_tensor(train_y, dtype = tf$float32)

  model <- create_model(ncol(train_x))

  history <- model %>% fit(
    x = train_x, y = train_y,
    epochs = 50,
    batch_size = 16,
    verbose = 0
  )

  test_predictions <- model %>% predict(test_x)
  neural_net_predictions[test_indices] <- test_predictions
}

neural_net_predictions <- 100 * (neural_net_predictions - min(neural_net_predictions)) / 
                          (max(neural_net_predictions) - min(neural_net_predictions))

cat("Neural Network predictions generated successfully.\n")
```

## Neural Network Plot

```{r}
plot_data <- data.frame(
  Actual = pca_data$over_under_2020_2023,
  Predicted = neural_net_predictions
)

ggplot(plot_data, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Actual vs. Predicted Values from Neural Network",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()

plot_data$residuals <- plot_data$Actual - plot_data$Predicted

ggplot(plot_data, aes(x = Actual, y = residuals)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals of Neural Network Predictions",
       x = "Actual Values",
       y = "Residuals") +
  theme_minimal()

library(ggplot2)

plot_data$GentrificationStatus <- ifelse(plot_data$Actual > 10, "Gentrified", "Non-Gentrified")

ggplot(plot_data, aes(x = Actual, y = Predicted, color = GentrificationStatus)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Gentrified" = "red", "Non-Gentrified" = "green")) +
  labs(
    title = "Scatter Plot of Neural Network Predictions vs Actual Values",
    x = "Percent change in Average Home Value Compared to State Average",
    y = "Predicted Gentrification Score (Normalized)",
    color = "Gentrification Status"
  ) +
  theme_minimal()
```

## Boundary classification code setup

```{r}
train_index <- sample(seq_len(nrow(pca_data)), size = 0.7 * nrow(pca_data))
train_data <- pca_data[train_index, ]
test_data <- pca_data[-train_index, ]

combined_indicator <- (pca_predictions + neural_net_predictions) / 2
pca_data$combined_indicator <- combined_indicator

pca_data$GentrificationStatus <- ifelse(pca_data$over_under_2020_2023 > 10, "Gentrified", "Non-Gentrified")
actual_labels <- ifelse(test_data$GentrificationStatus == "Gentrified", 1, 0)
```


```{r}
# summary(pca_data$combined_indicator)
table(pca_data$GentrificationStatus)

library(ggplot2)
ggplot(pca_data, aes(x = combined_indicator, fill = GentrificationStatus)) +
  geom_density(alpha = 0.7) +
  labs(title = "Distribution of Combined Indicator by Gentrification Status",
       x = "Combined Indicator",
       y = "Density")
```


```{r}
table(qda_predictions$class)
table(lda_predictions)
table(svm_predictions)
```

# County QDA Comparison

```{r}
library(MASS)

qda_model <- qda(GentrificationStatus ~ combined_indicator, 
                 data = train_data, 
                 prior = c(0.5, 0.5)) 


qda_predictions <- predict(qda_model, newdata = test_data)
predicted_labels_qda <- ifelse(qda_predictions$class == "Gentrified", 1, 0)
qda_mse <- mean((actual_labels - predicted_labels_qda)^2)

cat("QDA Mean Squared Error (MSE):", qda_mse, "\n")
```

## Combined QDA plots

```{r}
library(ggplot2)

grid <- data.frame(combined_indicator = seq(min(pca_data$combined_indicator), 
                                            max(pca_data$combined_indicator), 
                                            length.out = 200))

grid$prediction <- predict(qda_model, grid)$class

ggplot() +
  geom_density(data = train_data, aes(x = combined_indicator, fill = GentrificationStatus), alpha = 0.7) +
  geom_point(data = grid, aes(x = combined_indicator, y = 0, color = prediction), 
             size = 1.5, shape = "|") +
  labs(title = "QDA Decision Boundary with Combined Indicator",
       x = "Combined Indicator",
       y = "Density") +
  scale_color_manual(values = c("Gentrified" = "steelblue", "Non-Gentrified" = "lightgray")) +
  theme_minimal()
```

## LDA

```{r}
library(MASS)

lda_model <- lda(GentrificationStatus ~ combined_indicator, 
                 data = train_data, 
                 prior = c(0.5, 0.5))

lda_predictions <- predict(lda_model, newdata = test_data)$class
predicted_labels_lda <- ifelse(lda_predictions == "Gentrified", 1, 0)
lda_mse <- mean((actual_labels - predicted_labels_lda)^2)

cat("LDA Mean Squared Error (MSE):", lda_mse, "\n")
```

# Plot

```{r}
library(MASS)
library(ggplot2)

grid <- data.frame(combined_indicator = seq(min(pca_data$combined_indicator), 
                                            max(pca_data$combined_indicator), 
                                            length.out = 200))

grid$prediction <- predict(lda_model, grid)$class

ggplot() +
  geom_density(data = train_data, aes(x = combined_indicator, fill = GentrificationStatus), alpha = 0.7) +
  geom_point(data = grid, aes(x = combined_indicator, y = 0, color = prediction), 
             size = 1.5, shape = "|") +
  labs(title = "LDA Decision Boundary with Combined Indicator",
       x = "Combined Indicator",
       y = "Density") +
  scale_color_manual(values = c("Gentrified" = "steelblue", "Non-Gentrified" = "lightgray")) +
  theme_minimal()
```

## Error Rate

```{r}
lda_predictions <- predict(lda_model, newdata = pca_data)$class

actual_labels <- ifelse(pca_data$GentrificationStatus == "Gentrified", 1, 0)
predicted_labels <- ifelse(lda_predictions == "Gentrified", 1, 0)
mse <- mean((actual_labels - predicted_labels)^2)

cat("Mean Squared Error (MSE) for LDA classification:", mse, "\n")
```

## SVM

```{r}
library(e1071)
train_data$GentrificationStatus <- as.factor(train_data$GentrificationStatus)
test_data$GentrificationStatus <- as.factor(test_data$GentrificationStatus)

train_data$combined_indicator <- scale(train_data$combined_indicator)
test_data$combined_indicator <- scale(test_data$combined_indicator)

svm_model <- svm(GentrificationStatus ~ combined_indicator, 
                 data = train_data,
                 kernel = 'radial',
                 class.weights = c("Gentrified" = 3, "Non-Gentrified" = 1))

summary(svm_model)
```

# SVM with k-fold cross validation

``` {r}
library(e1071)

tuned_model <- tune(
    svm,
    GentrificationStatus ~ combined_indicator,
    data = train_data,
    kernel = "radial",
    ranges = list(cost = 10^(-2:2), gamma = 10^(-3:1))
)

best_model <- tuned_model$best.model
svm_predictions <- predict(best_model, newdata = test_data, probability = TRUE)

accuracy <- mean(svm_predictions == test_data$GentrificationStatus)
cat("Tuned SVM Accuracy:", accuracy, "\n")

conf_matrix <- table(Predicted = svm_predictions, Actual = test_data$GentrificationStatus)
print("Confusion Matrix:")
print(conf_matrix)
```

```{r}
library(ggplot2)

grid <- data.frame(combined_indicator = seq(min(train_data$combined_indicator),
                                            max(train_data$combined_indicator), 
                                            length.out = 500))

grid$prediction <- predict(svm_model, newdata = grid, probability = TRUE)
grid$normalized_prediction <- ifelse(grid$prediction == "Gentrified", 1, 0)

ggplot() +
  geom_density(data = train_data, aes(x = combined_indicator, fill = GentrificationStatus), 
               alpha = 0.7) +
  geom_point(data = grid, aes(x = combined_indicator, y = 0, color = prediction), 
             size = 1.5, shape = "|") +
  labs(title = "SVM Decision Boundary with Combined Indicator",
       x = "Combined Indicator",
       y = "Density") +
  scale_color_manual(values = c("Gentrified" = "steelblue", "Non-Gentrified" = "lightgray")) +
  scale_fill_manual(values = c("Gentrified" = "steelblue", "Non-Gentrified" = "lightgray")) +
  theme_minimal()
```

``` {r}
grid <- data.frame(combined_indicator = seq(min(train_data$combined_indicator),
                                            max(train_data$combined_indicator), 
                                            length.out = 1000))

svm_probs <- predict(svm_model, newdata = grid, probability = TRUE)
grid$decision_value <- attr(svm_probs, "decision.values")
boundary <- grid$combined_indicator[which.min(abs(grid$decision_value))]
print(paste("The decision boundary is at combined_indicator =", boundary))
```


```{r}
library(e1071)
library(ggplot2)

train_data$GentrificationStatus <- as.factor(train_data$GentrificationStatus)
train_data$combined_indicator <- scale(train_data$combined_indicator) 
test_data$combined_indicator <- scale(test_data$combined_indicator) 

svm_model <- svm(GentrificationStatus ~ combined_indicator, 
                 data = train_data,
                 kernel = 'radial')

# Predict decision boundary over a range of combined_indicator values
grid <- data.frame(combined_indicator = seq(min(train_data$combined_indicator, na.rm = TRUE),
                                            max(train_data$combined_indicator, na.rm = TRUE), length.out = 500))

grid$prediction <- predict(svm_model, newdata = grid)

# Create a plot of density and SVM decision boundary
ggplot() +
  # Density plots for gentrified and non-gentrified counties
  geom_density(data = train_data, 
               aes(x = combined_indicator, fill = GentrificationStatus), 
               alpha = 0.6) +
  # Decision boundary visualization using SVM model predictions
  geom_vline(xintercept = as.numeric(names(sort(table(grid$prediction), decreasing = TRUE))[2]),
             color = "red", linetype = "dashed", size = 1) +
  # Add plot labels
  labs(title = "SVM Decision Boundary and Combined Indicator Density",
       x = "Combined Indicator",
       y = "Density") +
  scale_fill_manual(values = c("Gentrified" = "blue", "Non-Gentrified" = "gray")) +
  theme_minimal() +
  theme(legend.position = "top")

```
``` {r}
library(e1071)
library(ggplot2)

train_data$combined_indicator <- scale(train_data$combined_indicator)
test_data$combined_indicator <- scale(test_data$combined_indicator)
summary(train_data$combined_indicator)
summary(test_data$combined_indicator)

tuned_model <- tune(
  svm,
  GentrificationStatus ~ combined_indicator,
  data = train_data,
  kernel = "radial",
  probability = TRUE,
  ranges = list(cost = 10^(-2:4), gamma = 10^(-4:2))
)
print(tuned_model)

best_model <- tuned_model$best.model
svm_raw_predictions <- predict(best_model, newdata = test_data)
table(svm_raw_predictions, test_data$GentrificationStatus)

grid <- data.frame(combined_indicator = seq(min(train_data$combined_indicator),
                                            max(train_data$combined_indicator),
                                            length.out = 500))

svm_predictions_prob <- attr(predict(best_model, newdata = grid, probability = TRUE), "probabilities")[, "Gentrified"]
grid$prob_prediction <- svm_predictions_prob
summary(grid$prob_prediction)

ggplot(grid, aes(x = combined_indicator, y = prob_prediction)) +
  geom_line(color = "blue", size = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "SVM Decision Boundary with Tuned Model",
       x = "Scaled Combined Indicator",
       y = "Predicted Probability of Gentrified") +
  theme_minimal()
```

``` {r}
library(cluster)
library(dplyr)

census_features <- census_data_2020 %>%
  dplyr::select(Vacancy_Rate_Percent_2020, Percent_White_Alone_2020, Percent_Black_Alone_2020, Percent_Below_Poverty_2020, Percent_Renter_Paying_30_2020)

census_features_ungr <- census_features %>%
  ungroup() %>%
  dplyr::select(-GEOID)  

scaled_features <- scale(census_features_ungr)
kmeans_result <- kmeans(scaled_features, centers = 3)
census_data_2020$Cluster <- kmeans_result$cluster
```


``` {r}
pca_result <- prcomp(scaled_features)

plot_data <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  cluster = as.factor(kmeans_result$cluster)  
)

library(ggplot2)
ggplot(plot_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2) +
  labs(title = "PCA Performed on K-Means Clusters",
       x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```

``` {r}
centers_df <- as.data.frame(kmeans_result$centers)
centers_df
```